Diversity and evolution of centromere repeats in the maize genome

### 2014-04-22 centc.tex
Ready for Kelly to comment

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage[utf8]{inputenc}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{natbib}
\usepackage{lscape}

@@

\jri{needs some basic setup. we start with alignment (?) and then do what to it? what was the code written in? is the code available? if so cite github url}.

We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8 with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  Therefore, we used simulations in python to count the number of sites with shared diversity across copies of CentC (code available on \url{GITHUB} \pb{we should talk about this.}).  Our simulation assumed that CentC has been evolving for the 1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$, and one generation per year.

@@

Fluorescent in situ hybridization (FISH) was used to compare abundance of CentC repeats in chromosomes of a maize x \emph{parviglumis} F1 hybrid and a maize x \emph{Z. luxurians} F1 hybrid. FISH protocols closely followed those of X and Y.
\jri{Need Kelly to fill in}

@@

We found a total of 12,162 CentC variants in the maize reference genome and the unassembled BACs.  Of these 12,162 copies, 8,259 were unique over their full length. No CentC occurred more than 10 times in the genome, and the vast majority ($>$75\%, Supplemental Table \ref{supp.occurence}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem, identical duplicates.  Genome-wide CentC locations also show that nearly all of the 10,639 CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Cluster size varied from single CENTC copies to 84KB, with a mean of ~7KB (approximately 45 CentC copies). Chromosomes varied greatly in copies of CentC, though we know that centromere assemblies for all of the chromosomes are not complete.   For example, in mapping reads from ChIP with CENH3 to an oat plant with a single maize chromosome (OMA) (\pb{whats the ref here? which paper did jiming send us reads from?}), we find reads from many chromosomes have hits to the unassembled BACs (Supplemental table \ref{gaby}).  In particular, chromosome 6 had many more reads aligning to the BACs than it did to its own centromere, suggesting a particularly lacking assembly.  Examining total repeat number, Chromosome 7 had the most CentC, with 3,200 copies, while chromosome 6 had the fewest with 32 copies.

@@

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}.  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Supplementary Fig Histogram \jri{add}).  While Schnable et al. \citet{Schnable2010} do not assign subgenome status to repetitive regions, we assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable); of the 17 clusters $>$20KB, 16 were unambiguously assigned to subgenome 1.  Even correcting for the genome-wide overrepresentation of subgenome 1 (62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10KB and 20KB clusters).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
  \includegraphics{circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  High confidence regions are colored with darker colors while low confidence regions are colored with lighter colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. \jri{fix legend. what is middle outer inner ring?}}
\label{circos}    
\end{figure}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{heatmap.png}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top) other copies within its island of tandem repeats delineated by dotted lines and (bottom) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map.  
}
\label{heatmap}    
\end{figure}

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{TW_CentCgraph.jpeg}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin for CentC's on the y-axis. The intersection shows the number of CentC's in the group found on that chromosome, scaled logarithmically.  CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown. \jri{low quality fig. use pdf not png or jpeg.}
}
\label{pcoa}    
\end{figure}

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Figure3_CentC_Abundance.png}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. \jri{center these figs}
}
\label{abundance}    
\end{figure}

@@

Shotgun sequence data from the maize HapMap v2 \citep{Chia2012}, reveals a significantly greater abundance of CentC in teosinte than in domesticated maize (p$<0.01$; Figure \ref{abundance}). Teosintes have more CentC than inbred maize.  Further support for differences between maize and its wild relatives comes from shotgun sequence data from \emph{Z. luxurians} \citep{Tenaillon2011}.  Analysis of these data find nearly twice as much CentC in \emph{Z. luxurians} as the maize inbred B73.  To corroborate these results, we performed fluorescent in-situ hybridization of F1 crosses between inbred maize and teosinte to determine if cytological observations agreed with our sequencing findings.  Cytology needed to be performed on an F1 cross so that we could compare relative probe fluorescence of the chromosomes within a single individual.  FISH data supports our observation that the teosintes \emph{parviglumis} and \emph{Z. luxurians} have more CentC than inbred maize (Figure \ref{abundance}).

@@

Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression of its genes than maize subgenome 2.  \jri{cite needed} Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be very common across plants (\emph{A. thaliana} \cite{Cheng2012} and \emph{Gossypium} Kapp and Wendel unpublished).  \pb{Two comments about this: first, its a terrible sentence so help make this better, because i think the point is really nice.  second, the best citations i found for cotton was an unpublished ref that is still not published from what i can gather.  advice? e.g. \url{http://www.plosone.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pone.0036442} also check out papers that cite Schabs: \url{http://scholar.google.com/scholar?cites=5313714612108479532&as_sdt=2005&sciodt=0,5&hl=en} } There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

@@

When studying CentC changes through domestication, our findings show that the repeat has experienced a decrease in copy number over time without a noticeable change in the structure of the repeat arrays. Using PacBio long read sequencing, we confirm genome-wide the observation we see on chromosomes 2 and 5, that most copies of CentC exist in large tandem arrays in all \emph{Zea} taxa.  \jri{i think the discussion should largely mirror the results. this doesn’t go in the domestication part really.}  \pb{the pacbio is no where in the results, just methods.  Ok here in discussion?} \jri{nope, need pacbio in results, even if briefly} From short read sequencing data, we show that modern maize has reduced genomic abundance of CentC when compared to the teosintes, a finding confirmed through FISH (Figure \ref{abundance}).  Both teosintes within the species \emph{mays} and its sister species \emph{luxurians} have elevated levels of CentC.  Lower levels of centromere repeats in inbred maize contrasts previous studies that characterized the changing abundance of other common repeats.  For example, the abundance of most transposable element families increased after domestication \citep{Chia2012}.  Knowing that TE abundance increases in domesticated maize, we might have expected CentC content to increase as well if centromere size had to expand alongside most other repetitive content.  Alternatively, due to their structural role in kinetichore formation, we might have also expected centromere repeats to be largely excluded from the genome wide fluctuations in repetitive content assuming that selection exists to maintain centromere size.  Instead, the decrease in CentC content may indicate an active process of removing CentC.  We hypothesize that the removal of CentC content may correlate with genome size, as the sharp decrease in knob content through domestication actually led to an overall smaller genome size in inbred maize. The correlation between  functional  centromere size and genome size has been observed in grass species \citep{Zhang2012}. \jri{the problem here is kelly shows functional cent. size, not repeat array size.   } \pb{i dont think this changes the point, CentC, whether in or out the kinetichore, seems to maintain the same relatedness/diversity, not sure this is addressed so left comments in here} Further investigation would be required to discover whether only those copies of CentC outside of the active kinetochore are being deleted.

\pb{Michelle is pretty adamant that this isnt a TE. As i found, no GAG or POL, but still a repeat at ends.  Strike the whole thing, or worthwhile mentioning something about it? } \jri{i say we keep in, but with a more nuanced explanation.  basically we can say it is tandem. i say some sleuthing may be in order -- simple blast hits find it in oryza and vitis telmore repeats...}We also identify the cent4 repeat as the LTR retrotransposon RLX\_sela.  Characterization of the TE consensus sequences shows that it contains a 232bp LTR on both ends but lacks any of the protein sequences for autonomous transposition, such as GAG and POL complexes.  Furthermore, RLX\_sela has several \jri{why brackets? is this uncerstain?}\pb{had brackets because its unclear how I want to group these and set lengths of them, so i settle for several.} regions showing $>60$bp of homology to the 180bp knob repeat. The lack of binding between CenH3 and RLX\_sela seen in the ChIP data further suggests that this repeat is likely pericentromeric rather and not involved in centromere function, a result echoed by  \citet{Jin2004}, who showed that Cent4 probes lagging behind CentC probes in cell division. Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  Given that the centromere on chromosome 4 shows signs of strong selection during domestication \citep{Hufford2012} \jri{doublecheck i'm not misremembering this please} \pb{Super high peak in landracevsimproved just below centromere, and reasonable peak around centromere in teovslandrace. However, I am pretty sure there are enough peaks to say sure, one of those is this}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

@@

\begin{acknowledgements}
\jri{add vince, anne lorant?}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Lauren Sagara, Gernot Presting,  NSF summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista, and Siddharth Bhadra-Lobo for helpful discussion.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage[utf8]{inputenc}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{float}

@@

We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8 with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  Therefore, we used simulations in python to count the number of sites with shared diversity across copies of CentC (code available on \url{GITHUB}).  Our simulation assumed that CentC has been evolving for the 1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$, and one generation per year.

@@

Fluorescent in situ hybridization (FISH) was used to compare abundance of CentC repeats in chromosomes of a maize x \emph{parviglumis} F1 hybrid and a maize x \emph{Z. luxurians} F1 hybrid. FISH protocols closely followed those of X and Y.
\pb{Kelly, would you mind helping fill this in?}

@@

We found a total of 12,162 CentC variants in the maize reference genome and the unassembled BACs.  Of these 12,162 copies, 8,259 were unique over their full length. No CentC occurred more than 10 times in the genome, and the vast majority ($>$75\%, Supplemental Table \ref{supp.occurence}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem, identical duplicates.  Genome-wide CentC locations also show that nearly all of the 10,639 CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Cluster size varied from single CENTC copies to 84KB, with a mean of ~7KB (approximately 45 CentC copies). Chromosomes varied greatly in copies of CentC, though we know that centromere assemblies for all of the chromosomes are not complete.   For example, in mapping reads from ChIP with CENH3 to an oat plant with a single maize chromosome (OMA) (\pb{whats the ref here? which paper did jiming send us reads from? Gaby?}), we find reads from many chromosomes have hits to the unassembled BACs (Supplemental table \ref{gaby}).  In particular, chromosome 6 had many more reads aligning to the BACs than it did to its own centromere, suggesting a particularly lacking assembly.  Examining total repeat number, Chromosome 7 had the most CentC, with 3,200 copies, while chromosome 6 had the fewest with 32 copies.

@@

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}.  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Supplementary Fig \ref{Supp_clusterhist).  While Schnable et al. \citet{Schnable2011} do not assign subgenome status to repetitive regions, we assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable); of the 17 clusters $>$20KB, 16 were unambiguously assigned to subgenome 1.  Even correcting for the genome-wide overrepresentation of subgenome 1 (62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10KB and 20KB clusters).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
  \includegraphics{circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  High confidence regions are colored with darker colors while low confidence regions are colored with lighter colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20KB in length.}
\label{circos}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{heatmap.png}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top) other copies within its island of tandem repeats delineated by dotted lines and (bottom) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map.  
}
\label{heatmap}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Rplot02}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin for CentC's on the y-axis. The intersection shows the number of CentC's in the group found on that chromosome, scaled logarithmically.  CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown. 
}
\label{pcoa}    
\end{figure}

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\centering
\includegraphics[width=1\textwidth]{Figure3_noedge}
\caption{(a) Mb of CentC in genomic libraries of maize and teosinte.  Box plots show data from \citet{Chia2012}.  Points show data for maize inbred B73 and the teosinte \emph{Z. luxurians} from \citet{Tenaillon2011}.  For comparison, the data point of maize inbred B73 in \citet{Chia2012} is shown with a tick mark on the box plot. (b) FISH images from an F1 hybrids between maize and the teosinte \emph{parviglumis}, and maize and the teosinte \emph{Z. luxurians}.  Upper row shows the karyotype for the teosinte and the lower row shows the maize karyotype.  Red=CentC, Green=CRM. 
}
\label{abundance}    
\end{figure}

@@

Shotgun sequence data from the maize HapMap v2 \citep{Chia2012}, reveals a significantly greater abundance of CentC in teosinte than in domesticated maize (p$<0.01$; Figure \ref{abundance}). Teosintes have more CentC than inbred maize.  Further support for differences between maize and its wild relatives comes from shotgun sequence data from \emph{Z. luxurians} \citep{Tenaillon2011}.  Analysis of these data find nearly twice as much CentC in \emph{Z. luxurians} as the maize inbred B73.  To corroborate these results, we performed fluorescent in-situ hybridization of F1 crosses between inbred maize and teosinte to determine if cytological observations agreed with our sequencing findings.  Cytology needed to be performed on an F1 cross so that we could compare relative probe fluorescence of the chromosomes within a single individual.  FISH data supports our observation that the teosintes \emph{parviglumis} and \emph{Z. luxurians} have more CentC than inbred maize (Figure \ref{abundance}).  Using whole genome shotgun PacBio long reads, we further investigated the overall structure of repeats across the different \emph{Zea} species.  Percentages of the libraries showing tandem repeats did not differ greatly across the taxa (Supplementary Figure \ref{supp.pacbio}).

@@

Our study traces the changes in centromere repeat genetic relatedness across maize chromosomes to show how ancient tetraploidy and subsequent evolution has impacted centromeres.  Most interestingly, we show that most large arrays of CentC in the maize reference genome derive from maize1 subgenome, which is known to have lower gene loss and higher average expression of its genes than maize subgenome 2 \cite{Schnable2011}.  Regulatory microRNA’s are known to correspond to centromeric repeats \citep{ReinhartBartel2002}.  Therefore, it may be worthwhile to further investigate the potential correlation between centromere retention and gene expression as the phenomenon of higher expression of genes from one ancestor may be very common across plants (\emph{A. thaliana} \cite{Cheng2012} and \emph{Gossypium} Kapp and Wendel unpublished).  \pb{Two comments about this: first, its a terrible sentence so help make this better, because i think the point is really nice.  second, the best citations i found for cotton was an unpublished ref that is still not published from what i can gather.  advice? e.g. \url{http://www.plosone.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pone.0036442} also check out papers that cite Schabs: \url{http://scholar.google.com/scholar?cites=5313714612108479532&as_sdt=2005&sciodt=0,5&hl=en} } There are several possible explanations for the differential contribution of subgenome 1 to the diversity of CentC in the extant maize  genome. We put forth several hypotheses that may explain the strong bias for CentC to be located within a block from subgenome 1. Subgenome 2 may have had severely reduced quantities of CentC and therefore never contributed equally.  The CentC from subgenome 2 may have decayed rapidly and thus less of it remains.  Lastly, the CentC from subgenome 1 may have expanded in copy number since the allopolyploidy event.  Though we are unable to conclusively identify the reason, the lack of a large number of identical tandem duplicates  and lack of close proximity highest genetic relatedness pairs suggests that any large scale expansion of CentC has not been recent.  Also, CentC’s in subgenome 2 cluster do not appear to have accumulated excess mutation, suggesting that they are not decaying more rapidly than their subgenome 1 counterparts.  We therefore believe that it is most likely that the subgenomes had an unequal contribution of CentC at the formation of the allopolyploid, though more research about the ancient parents is required.

@@

When studying CentC changes through domestication, our findings show that the repeat has experienced a decrease in copy number over time without a noticeable change in the structure of the repeat arrays. Using PacBio long read sequencing, we confirm genome-wide the observation we see on chromosomes 2 and 5, that most copies of CentC exist in large tandem arrays in all \emph{Zea} taxa.  From short read sequencing data, we show that modern maize has reduced genomic abundance of CentC when compared to the teosintes, a finding confirmed through FISH (Figure \ref{abundance}).  Both teosintes within the species \emph{mays} and its sister species \emph{luxurians} have elevated levels of CentC.  Lower levels of centromere repeats in inbred maize contrasts previous studies that characterized the changing abundance of other common repeats.  For example, the abundance of most transposable element families increased after domestication \citep{Chia2012}.  Knowing that TE abundance increases in domesticated maize, we might have expected CentC content to increase as well if centromere size had to expand alongside most other repetitive content.  Alternatively, due to their structural role in kinetichore formation, we might have also expected centromere repeats to be largely excluded from the genome wide fluctuations in repetitive content assuming that selection exists to maintain centromere size.  Instead, the decrease in CentC content may indicate an active process of removing CentC.  We hypothesize that the removal of CentC content may correlate with genome size, as the sharp decrease in knob content through domestication actually led to an overall smaller genome size in inbred maize. The correlation between  functional  centromere size and genome size has been observed in grass species \citep{Zhang2012}.  Further investigation would be required to discover whether only those copies of CentC outside of the active kinetochore are being deleted.

We also sought to further characterize the chromosome specific repeat cent4 \citep{Page2001}.  Previous work showed that Cent4 probes lagged behind CentC probes in cell division \citep{Jiang2002}, suggesting that the repeat is not involved in the active kinetochore and rather located in the pericentromere.  A lack of enrichment of cent4 in ChIP data agrees with a pericentromeric location.  Using the published cent4 sequences, we identify the cent4 repeat as the LTR retrotransposon RLX\_sela in the transposable element database \citep{Baucom2009}.  Characterization of the TE consensus sequence shows that it contains repetitive motifs with homology to both knob and telomere repeats.  It lacks any of the protein sequences necessary for autonomous transposition, such as GAG and POL complexes.  Previous work in rice has documented the presence of nonautonomous LTR retrotransposons  in or near the centromere \citep{Jiang2002}.  However, the RLX\_sela sequence also appears to be missing some of the necessary primer binding sites that would distinguish it as a nonautonomous TE, suggesting that it may be a tandem repeat, and further sequencing of the region will be required to uncover the role of cent4.   Given that the centromere and pericentromere on chromosome 4 show signs of strong selection during domestication \citep{Hufford2012}, it is plausible that the increase in abundance of Cent4 seen in domesticated maize is due to the effects of linked selection on a locus important for domestication rather than novel insertions or duplications.

@@

\begin{acknowledgements}
We wish to thank Pacific Biosciences for sequencing resources.  We thank Vince Buffalo, Lauren Sagara, Anne Lorant, Michelle Stitzer, Gernot Presting,  NSF summer exchange program interns Cesar Alvarez Mejia, Aurelio Hernandez Bautista, and Siddharth Bhadra-Lobo for helpful discussion.

@@

\begin{figure}[h!]
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Supp_clusterhist}
\setcounter{figure}{0}
\caption{CentC cluster size across all chromosomes.}
\label{Supp_clusterhist}    
\end{figure}
